Analysis \& Visualization of Call Center Data using Python

### 2018-08-28 paper.tex
Restructured paper, new plots being uploaded to github

An analysis of call volume by month was conducted using the daily data, which provides aggregate totals by day for a 2-year period. The relevant data was manipulated by taking the mean of the daily call volume for every record in each month.  Upon completion, the data was then plotted in a bar chart format, allowing the user to quickly identify any differences.  This chart can be viewed in Figure 1.

As an exercise in thoroughness, these findings were subjected to a two-sided t-test for statistical significance.  The daily call volumes for each month were isolated and stored as lists within a Python dictionary.  The scipy.stats package was used to conduct the test.  For the purposes of this test, the null hypothesis stated that ``the two months have identical expected values."  For the null hypothesis to be disproved, a standard threshold p-value of 0.05 was used.  The heatmap in Figure 2 shows the p-value between each pairing of months.  From this information, the city can see which months are significantly busier/quieter than others.  By subjecting the data to such a test, important points can be quickly identified that may or may not be obvious in Figure 1.

\begin{figure}[h]
	\includegraphics[scale=.65]{Heatmap.png}
	\caption{P-values resulting from t-test on daily call volume by month}
\end{figure}

@@

  \begin{Verbatim}[fontsize=\small]

def string_to_sec(df, col_name):
    ''' 
    This function accepts a dataframe and 
    column name (in quotes)
    It converts the column from XX:XX:XX to
     seconds and returns a list.
    Pandas must be imported and 
    aliased as pd
    '''
    new = []
    for i in range(len(df[col_name])):
        x = df.loc[i][col_name]
        if x.count(":") == 1:
            ''' For Min:sec '''
            y, z = x.split(":")
            y = pd.to_numeric(y)
            z = pd.to_numeric(z)
            q = (y * 60) + z
        elif x.count(":") == 2:
       '''For H/Min:Min/sec:Sec/Nothing '''
            h, y, z = x.split(":")
            if len(h) == 2:
                ''' This assumes if the
               first number is 2 digits,
               it is minutes'''
                h = pd.to_numeric(h)
                y = pd.to_numeric(y)
                q = (h * 60) + y
             elif len(h) == 1:
              '''Assuming first digit is 
                single, means hours'''
                h = pd.to_numeric(h)
                y = pd.to_numeric(y)
                z = pd.to_numeric(z)
                q = (h*3600) + (y*60) + z
             else:
                 q = np.NaN
         else:
              q = np.NaN
         new.append(q)
  return new
  \end{Verbatim}

This function returns a list, which is then evaluated for reasonableness and appended to the dataframe.  By doing this, every call record now has a new attribute for duration in seconds.

Calls by department were then analyzed and visualized by average duration and total volume.  As expected, the departments' respective ordering in each analysis differed.  As a more direct approach to the measure of total call volume, the total departmental call times were calculated.  Of all the approaches, this calculation of total call time may be the most helpful to the city, as this can directly show where time and budget is being spent.  This visualization can be seen in the pie chart in Figure 3.

\begin{figure}[h]
  \includegraphics[scale=0.3]{piechart.png}
  \caption{A visual of total volume by department}
 \end{figure}

@@

Analyzing calls by topic proved to be one of the most challenging and interesting aspects of this analysis.  The topic data was used again here due to its recentness and the homogeneity of topic format.  Among this dataset, 426 unique call topics existed.  The data was then processed to show total call count and average call duration by topics.  This data was then merged with a separate dataframe, indexed by topic with department as the only attribute.  Once completed, the resulting dataframe contained topic, count, average call duration, and corresponding department.  At this point, the data was ready for analysis.

\subsection{Duration}
Call duration by topics was calculated in the dataframe creation above, using the ``.groupby()'' method in Pandas.  However, when the dataframe was sorted to see the quickest and longest average duration, several topics with a low volume rose to the top, some with only one total call.  In a sense, such records can be viewed as outliers, and were thus removed from reporting in the final dashboard.  In the end, only topics with a total volume above the first quartile (of total calls) were included in this analysis.  By reducing the topics analyzed, the call center will be able to focus on topics that take a notable duration to resolve, but that also occur frequently enough to warrant attention.  This code, however, can be easily adjusted to a different threshold, if management so desires.

\subsection{Volume}
The total number of calls by topic was then analyzed.  The dataframe that was used in the previous analysis was again used for this.  Using the ``.groupby()" method in Pandas in conjunction with ``.count()", the total number of calls for each topic was appended to the dataframe.  The most frequent call topics were quickly identified, and many showed a significant overlap with the highest volume departments.  As a result, the volume by topic in and of itself would not serve much use to the city.  Therefore, the top 7 highest volume departments (identified in Figure 2) were isolated from the rest of the data.  The individual topics within these departments were then identified.  By approaching the analysis in this manner, the city can easily see the most popular topics in the busiest departments.  In order to visualize this data in the most effective manner, a jitterplot was incorporated into the final dashboard.  This plot can be seen in Figure 4.

\begin{figure}[h]
  \includegraphics[scale=0.3]{jitterplot.png}
  \caption{Jitterplot of call volume by topic for the busiest departments.  Hovertool is utilized to show topic.  The topic in diamond is not plotted to scale}
\end{figure}

>>>>

An analysis of call volume by month was conducted using the daily data, which provides aggregate totals by day for a 2-year period. The relevant data was manipulated by taking the mean of the daily call volume for every record in each month.  As an exercise in thoroughness, these findings were subjected to a two-sided t-test for statistical significance.  The daily call volumes for each month were isolated and stored as lists within a Python dictionary.  The scipy.stats package was used to conduct the test.  For the purposes of this test, the null hypothesis stated that ``the two months have identical expected values."  For the null hypothesis to be disproved, a standard threshold p-value of 0.05 was used.

	\subsection{Calls by Day of Week}

Calls by day of week were able to be obtained from two data files:  daily data and topic data.  The first conveniently had an attribute consisting of an integer corresponding to a weekday.  The topic data included a date of call, but it did not have a direct day attribute.  Such an attribute was able to be created through the use of the ``to\_datetime()" and ``weekday()" functions in Pandas.  Once this attribute was created in the same format as the daily data, the dataframes were concatonated using ``concat()".  Aggregate call totals by day were then simply calculated using the ``.count()" method in Pandas.

@@

  \begin{Verbatim}[fontsize=\small]
 def string_to_sec(df, col_name):
     ''' 
     This function accepts a dataframe and 
     column name (in quotes)
     It converts the column from XX:XX:XX to
     seconds and returns a list.
     Pandas must be imported and 
     aliased as pd
     '''
     new = []
     for i in range(len(df[col_name])):
         x = df.loc[i][col_name]
         if x.count(":") == 1:
            ''' For Min:sec '''
             y, z = x.split(":")
             y = pd.to_numeric(y)
             z = pd.to_numeric(z)
             q = (y * 60) + z
         elif x.count(":") == 2:
          '''For H/Min:Min/sec:Sec/Nothing '''
             h, y, z = x.split(":")
            if len(h) == 2:
             ''' This assumes if the
               first number is 2 digits,
               it is minutes'''
                h = pd.to_numeric(h)
                y = pd.to_numeric(y)
                q = (h * 60) + y
             elif len(h) == 1:
              '''Assuming first digit is 
                single, means hours'''
                h = pd.to_numeric(h)
                y = pd.to_numeric(y)
                z = pd.to_numeric(z)
                q = (h*3600) + (y*60) + z
             else:
                q = np.NaN
         else:
              q = np.NaN
         new.append(q)
  return new
  \end{Verbatim}

This function returns a list, which is then evaluated for reasonableness and appended to the dataframe.  By doing this, every call record now had a new attribute for duration in seconds.

@@

Analyzing calls by topic proved to be one of the most challenging aspects of this analysis.  The topic data was used again here due to its recentness and the homogeneity of topic format.  Among this dataset, 426 unique call topics existed.  The data was then processed to show total call count and average call duration by topics.  This data was then merged with a separate dataframe, indexed by topic with department as the only attribute.  Once completed, the resulting dataframe contained topic, count, average call duration, and corresponding department.  At this point, the data was prepared for analysis.
\par
Call duration by topics was calculated in the dataframe creation above, using the ``.groupby()'' method in Pandas.  However, when the dataframe was sorted to see the quickest and longest average duration, several topics with a low volume rose to the top, some with only one total call.  In a sense, such records can be viewed as outliers, and were thus removed from reporting in the final dashboard.  In the end, only topics with a total volume above the first quartile (of total calls) were included in this analysis.  By reducing the topics analyzed, the call center will be able to focus on topics that take a notable duration to resolve, but that also occur frequently enough to warrant attention.  This code, however, can be easily adjusted to a different threshold, if management so desires.
\par
The total number of calls by topic was then analyzed.  The dataframe that was used in the previous analysis was again used for this.  Using the ``.groupby()" method in Pandas in conjunction with ``.count()", the total number of calls for each topic was appended to the dataframe.  The most frequent call topics were quickly identified, and many showed a significant overlap with the highest volume departments.  As a result, the volume by topic in and of itself would not serve much use to the city.  Therefore, the top 7 highest volume departments (identified in Figure 2) were isolated from the rest of the data.  The individual topics within these departments were then identified.  By approaching the analysis in this manner, the city can easily see the most popular topics in the busiest departments.

Upon analysis, the months with the highest average daily call volumes were July, April, Jun, and May (in that order).  The results can be viewed in Figure 1.  The standard errors for each month (also plotted in Figure 1) do not appear to be so large as to discredit the significance of these differences.  March, interestingly, has the lowest average volume by a large margin.  This is especially intriguing since it is immediately prior to what can be seen as the ``busy season".

The heatmap in Figure 2 shows the resulting p-value between each pairing of months.  From this information, the city can see which months are significantly busier/quieter than others.  By subjecting the data to such a test, important points can be quickly identified that may or may not be obvious in Figure 1.  For example, June's call volumes are not statistically different from those in April.  March, however, has a statistically lower daily average than every other month.

\begin{figure}[h]
	\includegraphics[scale=.5]{Heatmap.png}
	\caption{P-values resulting from t-test on daily call volume by month}
\end{figure}

The analysis of calls by day of week incorporated over 588,000 calls into the results.  The output, therefore, can be viewed as information-rich and an excellent measure of call center activity, since the combined data sources span from 2013 into 2018.  Given the variablity of the vast amount of data, an interactive historgram was chosen as the best method for communication of results.  The user can click to show/hide days of the week, and the chart will layer the call distributions over each other for comparison.  Figure 3 shows this graph, with Monday and Thursday visible.  The general distribution of each day is similar and appears to follow a 
%what the hell kind
distribution.  Monday has the highest volume distribution, while Friday has the lowest.  This is also evidenced in a swarm plot of only the daily data in Figure 4.

\begin{figure}[h]
	\includegraphics[scale=.2]{daily_calls.png}
	\caption{Histogram of total daily calls by day of week, 2013-2018}
\end{figure}

\begin{figure}[h]
	\includegraphics[scale=.6]{swarm.png}
	\caption{Swarmplot of total daily calls by day of week, 2013-2015}
\end{figure}

%analysis here
Calls by department were analyzed by two measures:  average duration and total volume.  As expected, the results of each analysis differed.  Departments such as Solid Waste and Water Works had the highest call volumes, while Code Enforcement and the Mayor's Office has the longest average durations.  Given this incongruency, it is not immediately clear where the call center's resources are being focused.  As a more direct approach to the measure of total call volume, the total departmental call times were calculated.  Of all the approaches, this calculation of total call time may be the most helpful to the city, as this can directly show where time and budget is being spent.  This visualization can be seen in the pie chart in Figure 5.

\begin{figure}[h]
  \includegraphics[scale=0.25]{piechart.png}
  \caption{A visual of total volume by department}
 \end{figure}

\subsection{Duration}

\subsection{Volume}

In order to visualize this data in the most effective manner, a jitterplot was incorporated into the final dashboard.  This plot can be seen in Figure 6.

\begin{figure}[h]
  \includegraphics[scale=0.25]{jitterplot.png}
  \caption{Jitterplot of call volume by topic for the busiest departments.  Hovertool is utilized to show topic.  The topic in diamond is not plotted to scale}
\end{figure}
